---
title: "Prediction on U.S. 2020 Presidential Election"
author: "Boyuan Cui, Zhaocheng Li"
date: "November 2, 2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(knitr)

# Loading in the cleaned survey Data
survey_data <- read_csv("outputs/survey_data.csv")

# Loading in the cleaned census Data
census_data <- read_csv("outputs/census_data.csv")

```

# Abstract

__Key Words:__ U.S. 2020 Election, Popular Vote, Post-stratification, Logistic Regression

This report is an analysis to predict the U.S. 2020 federal election. The analysis employs the post-stratification technique by dividing the data into 838,113 cells according to different age, gender, state, race, education, and income. The final result of popular vote proportion prediction is 39.27% for Biden and 38.57% for Trump, and the electoral vote is 301 for Biden and 237 for Trump.

# Introduction

The U.S. presidential election has always been a popular topic of the world for every four year, and as the election day comes closer, we are all looking forward to see who will become the next U.S. president. This report is interested in predicting the U.S. 2020 election popular vote for each candidate (Joe Biden and Donald Trump). We will use the post-stratification technique in our analysis.

# Data

We extract the survey data from Democracy Fund + UCLA Nationscape [1] and the census data from IPUMS USA [2]. We remove all the examples with NA data from these two datasets. 

The survey data is collected on June 25, 2020 [1], and the census data is collected in 2018 [2].

# Model

## Model Specifics

We choose the logistic regression model to estimate the outcome of the presidential election. Because we are trying to find out whether the voter will vote for Trump or Biden in the 2020 election, the outcome is binary, so the logistic regression model is more appropriate and suitable in this case. 

We are using six predictor variables in our model, which are age, race, gender, income level and education level. These are the representative information in both census and survey dataset. To express the model mathematically, we have:

$$ \log(\frac{p}{1-p}) = \beta_0+\beta_1  x_{age} + \beta_2 x_{male} + \beta_3 x_{state} + \beta_4 x_{race} + \beta_5 x_{education} + \beta_6 x_{income} + \epsilon$$

where $p$ represents the proportion of voters who will vote for Trump/Biden, depending on which model we are using. Value of $\epsilon$ represents the residual of our fit. 

Parameter $\beta_0$ represents the intercept of the model, which, in this case, indicates the value of logit function $\log(\frac{p}{1-p})$ for a 0 year old Alabama female with 3rd grade or less education level and less than $15000 income.

Parameter $\beta_1$ represents the change in logit value by increasing age by 1.

Parameter $\beta_2$ represents the change in logit if the example is male. $x_{male}=1$ if subject is male, 0 if subject is female. No other possible values

Parameter $\beta_3$ represents the impact of state on the logit value. We use Alabama as a reference, so all change in logit is comparing with Alabama

Parameter $\beta_4$ represents the impact of race on the logit value. We use white as the reference.

Parameter $\beta_5$ represents the change in logit value by increasing education level by 1. Reference table of education level can be found at Appendix Table 3.

Parameter $\beta_6$ represents the change in logit value by increasing income level by 1. Reference table of income level can be found at Appendix Table 4.

Finally, we will fit two models, one for Trump and one for Biden, and we will compare the result to predict the winner of 2020 US presidential election.

Detail of the fitted model can be found at Appendix Table 5 and 6
```{r, include=FALSE}
survey_data = within(survey_data, race <- relevel(as.factor(race), ref = 7)) # make Single marital status as reference

# Creating the Model
model_trump <- glm(vote_trump ~ age + gender + state + race + education_level + income_level, 
            data=survey_data, family= "binomial")

# Model Results (to Report in Results section)
# summary(model)
# OR
 
 # Creating the Model
model_biden <- glm(vote_biden ~ age + gender + state + race + education_level + income_level, 
            data=survey_data, family= "binomial")
            
            
            
model_biden_no_state <- glm(vote_biden ~ age + gender + race + education_level + income_level,  data=survey_data, family= "binomial")
model_trump_no_state  <- glm(vote_trump ~ age + gender + race + education_level + income_level,   data=survey_data, family= "binomial")

# Model Results (to Report in Results section)
# summary(model)
# OR
```

## Post-Stratification

To estimate the popular vote for Biden and Trump, we will employ a post-stratification analysis.

To use the post-stratification technique, we will first divide the data into cells based on different age, gender, state, race, education, and income.

The reason why we choose these variables is that, we think these factors(age, gender,etc) are the basic voter's information which are being included in both census and survey dataset, so it might be easier to have access to these factors. Besides, the USA is a melting pot that consists of immigrants and people with various culture background from all over the world, so race plays an important role in the election and it should be considered into the analysis. Finally and the most importantly, historically, all the states in America could be classified into 3 categories: blue states, red states and swing states. The majority voters in blue states will choose Democratic party, while red states support Republican party. Two parties normally have similar amount of votes in swing states. Therefore, we think that state is also a crucial and interesting factor to determine the result of election; hence, we include it in the analysis.

Next, we apply the model fitted above to each cell to estimate the popular vote proportion for each candidate. To finish up our post-stratification analysis, we will calculate the weighted mean $\hat{y}^{PS}$ of all cells for each candidate with the formula below:

$$\hat{y}^{PS}=\frac{\sum_j N_j\cdot \hat{y_j} }{\sum_j N_j}$$

where $N_j,\hat{y_j}$ are the total count and model estimate of each cell, respectively.
  
The result can be found at Table 1 in the Result section.
```{r include=FALSE}

# Here I will perform the post-stratification calculation
# Post-stratification for Trump
# Post-stratification for Trump
census_data$logodds_estimate_T<-
  model_trump %>%
  predict(newdata = census_data)

census_data$estimate_T <-
  exp(census_data$logodds_estimate_T)/(1+exp(census_data$logodds_estimate_T))

trump = census_data %>%
  mutate(alp_predict_prop = estimate_T*n) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n))
census_data = mutate(census_data, alp_predict_prop = estimate_T*n)

# Post-stratification for Biden

census_data$logodds_estimate_B <-
  model_biden %>%
  predict(newdata = census_data)

census_data$estimate_B <-
  exp(census_data$logodds_estimate_B)/(1+exp(census_data$logodds_estimate_B))

biden = census_data %>%
  mutate(alp_predict_prop = estimate_B*n) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n))





# Post-stratification for Trump no state
census_data$logodds_estimate_TS<-
  model_trump_no_state %>%
  predict(newdata = census_data)

census_data$estimate_TS <-
  exp(census_data$logodds_estimate_TS)/(1+exp(census_data$logodds_estimate_TS))

trumps = census_data %>%
  mutate(alp_predict_prop = estimate_TS*n) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n))
census_data = mutate(census_data, alp_predict_prop = estimate_TS*n)

# Post-stratification for Biden no state

census_data$logodds_estimate_BS <-
  model_biden_no_state %>%
  predict(newdata = census_data)

census_data$estimate_BS <-
  exp(census_data$logodds_estimate_BS)/(1+exp(census_data$logodds_estimate_BS))

biden = census_data %>%
  mutate(alp_predict_prop = estimate_BS*n) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n), .groups='drop')

# state 
state_Trump = census_data %>%
  mutate(alp_predict_prop = estimate_TS*n) %>% group_by(state) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n))

state_Biden = data.frame ( census_data %>%
  mutate(alp_predict_prop = estimate_BS*n) %>% group_by(state) %>%
  summarise(alp_predict = sum(alp_predict_prop)/sum(n)), .groups='drop')




```


## Additional Analysis

In addition to the U.S. national popular vote, We also perform a post-stratification analysis by state. We first remove the variable state from the model and employ the same post-stratification technique on the new model. For the result, instead of two numbers of popular vote proportion in the previous analysis, we have predicted the popular vote proportion of each candidate for every state, which is illustrated by Fig. 1, 2, 3 in Appendix. 

Then we apply the winner-takes-all rule of the U.S. election system by distributing all the state's electoral vote to the candidate with higher popular vote at that state (except for Main and Nebraska which use proportional vote distribution rule) [3]. Finally we add up all the electoral votes for each candidate. The result is illustrated at Table 2 in the Result section.


# Results

```{r}

data.frame('Biden'=biden$alp_predict*100, 'Trump'=trump$alp_predict*100) %>% kable(caption = "Total Popular Vote(%)")

```

```{r}
voters <- c(9,3,11,6,55,9,7,3,3,29,16,4,4,20,11,6,6,8,8,4,10,11,16,10,6,10,3,5,6,4,14,5,29,15,3,18,7,7,20,4,9,3,11,38,6,3,13,12,5,10,3)

electoral <- data.frame(state=state_Trump$state, trump=rep(0, 51), biden=rep(0, 51))

for (i in 1:51){
sta = state_Trump$state[i]
elector = voters[i]

if (sta == 'maine' | sta == 'nebraska'){
  joe = (elector * state_Biden$alp_predict[i]) %>% round()
  donald = (elector * state_Trump$alp_predict[i]) %>% round()
  
  if ((joe + donald)!=elector){
    if (state_Biden$alp_predict[i] > state_Trump$alp_predict[i]){
      electoral$biden[i] <- joe + 1
      electoral$trump[i] <- donald
    } else {
      electoral$trump[i] <- donald + 1
      electoral$biden[i] <- joe
    }
  } else {
      electoral$biden[i] <- joe
      electoral$trump[i] <- donald
  }
  
} else {
  if (state_Biden$alp_predict[i] > state_Trump$alp_predict[i]){
    electoral$biden[i] <- elector
  } else {
    electoral$trump[i] <- elector
  }
}
  
}

data.frame(Biden=electoral$biden %>% sum(), Trump=electoral$trump %>% sum()) %>% kable(caption = 'Total Electoral Vote')
```

We estimate the popular vote proportion for Donald Trump is 38.567% and that for Joe Biden is 39.269%. This is based off our post-stratification analysis modelled by logistic regression, which accounted for age, gender, race, state, income, education.

For our additional analysis, we estimate the electoral vote for Donald Trump is 237 and that for Joe Biden is 301. This is based off our post-stratification analysis of states modelled by logistic regression, which accounted for age, gender, race, income, education. Detailed prediction of each state's electoral vote can be found at Table 7 in Appendix.

# Discussion

## Summary

In our report, we employ the post-stratification technique to predict the popular vote proportion of Biden and Trump. We first clean the census data by splitting the examples into 838,113 cells based on their age, gender, state, race, education level, and income level, which are our explanatory variables. Next we build two logistic regression models with our explanatory variables and use them to estimate the popular vote proportion of Biden and Trump for each cell. Ultimately, with each cell's estimate, we calculate the weighted mean of these estimates, which is our predicted popular vote of Biden and Trump.

We also conduct an analysis on post-stratification by state to find out the electoral vote of each candidate. The result is illustrated by graph Fig. 1, 2, 3 in Appendix.

## Conclusion

We predict that the popular vote in 2020 election is 39.269% for Joe Biden and 38.567% for Donald Trump. Although Biden has a higher popular vote, it does not necessarily mean that he is going to win, because in the US election system, the popular vote does not determine the winner, and actually candidate with less popular vote has become the president several times in the history [3]. 

What actually determines the winner is the electoral vote. Each state has certain number of electoral votes, and if one candidate has more state popular vote than the other, the leading candidate takes all the electoral vote of that state [3]. Therefore, the result in Table 2 determines the winner. As we can see, Joe Biden has more electoral votes than Donald Trump has (301 > 237), so we predict that Joe Biden will win the 2020 presidential election.

## Weakness

One of the major drawback of our analysis is that the survey data is outdated. The survey is conducted at June 25, 2020, but the election day is at November 3, 2020. Multiple major events had happened during this period, such as Donald Trump getting infected by COVID-19 and two presidential debates. These events could potentially change American people's opinion and hence affect the vote result.

The other weakness would be the small cell counts. When we divide the data into cells, there are many cells with small counts. These cells could bring uncertainties to the prediction and hence become problematic.

## Next Steps

Now that we have analyzes the popular vote through a frequentist approach, we can complicate the model to minimize the uncertainty brought by the small cell counts. One of the common way to deal with uncertainties is the Bayesian approach. We may consider to build a Bayesian logistic model for deeper analysis.

# References

[1] Tausanovitch, Chris and Lynn Vavreck. 2020. Democracy Fund + UCLA Nationscape, October 10-17, 2019 (version 20200814). Retrieved from https://www.voterstudygroup.org/publication/nationscape-data-set.

[2] Steven Ruggles, Sarah Flood, Ronald Goeken, Josiah Grover, Erin Meyer, Jose Pacas and Matthew Sobek. IPUMS USA: Version 10.0 [dataset]. Minneapolis, MN: IPUMS, 2020. https://doi.org/10.18128/D010.V10.0

[3] Presidential Election Process. (2020). In USA.gov. Retrieved November 2, 2020, from https://www.usa.gov/election

# Appendix

```{r}

state_Trump[1:17,] %>%
ggplot(aes(x =alp_predict, y = state, color="Trump")) + 
geom_point() +

xlab("2020 Popular Vote Predict") +
ylab("State") +
geom_point(data = state_Biden[1:17,], aes(alp_predict, state, color="Biden")) +
  scale_color_manual(values = c("blue","red")) + ggtitle("Fig.1 Popular Vote of State Alabama-Kansas (Alphabetically)")


state_Trump[18:34,] %>%
ggplot(aes(x =alp_predict, y = state, color="Trump")) + 
geom_point() +

xlab("2020 Popular Vote Predict") +
ylab("State") +
geom_point(data = state_Biden[18:34,], aes(alp_predict, state, color="Biden")) +
  scale_color_manual(values = c("blue","red")) + ggtitle("Fig.2 Popular Vote of State Kentucky-North Carolina (Alphabetically)")


state_Trump[35:51,] %>%
ggplot(aes(x =alp_predict, y = state, color="Trump")) + 
geom_point() +

xlab("2020 Popular Vote Predict") +
ylab("State") +
geom_point(data = state_Biden[35:51,], aes(alp_predict, state, color="Biden")) +
  scale_color_manual(values = c("blue","red")) + ggtitle("Fig.3 Popular Vote of State North Dakota-Wyoming (Alphabetically)")

edu <- c("3rd Grade or less",
        "Middle School - Grades 4 - 8",
        "Completed some high school ",
        "High school graduate",
        "Other post high school vocational training",
        "Completed some college, but no degree",
        "Associate Degree",
        "College Degree (such as B.A., B.S.)",
        "Completed some graduate, but no degree",
        "Masters degree",
        "Doctorate degree")

df <- data.frame(Education=edu, Level=c(0:10))
df[order(df$Level),] %>% kable(caption="Educational Level Reference")

df <- data.frame(Income=survey_data$household_income, Level=survey_data$income_level) %>% group_by(Income) %>% summarise(Level = mean(Level), .groups='drop')
df[order(df$Level),] %>% kable(caption="Income Level Reference")

broom::tidy(model_biden) %>% kable(caption="Popular Vote Logistic Model for Biden")
broom::tidy(model_trump) %>% kable(caption="Popular Vote Logistic Model for Trump")

electoral %>% rename(State=state, Trump=trump, Biden=biden) %>% kable(caption = "Electoral Vote of Each State")
```


